#### Working with columns

R is great at working with vectors because they are the fundamental data structure of the language. Since data frame columns are vectors, they are typically easy to operate on.

Basic statistical summary functions aren't too different from what you'd do in Excel.

```{r SummarizingData}
#to calculate the mean of the 'age' column in the original dataframe:
mean(tree_data$Age_yr)

#or to calculate the mean of the DBH vector we created:
dbh <- tree_data$DBH_in
mean(dbh)

#like-wise for the median, standard deviation, and minimum:
median(tree_data$Age_yr)
sd(tree_data$Age_yr)
min(tree_data$Age_yr)
```

Another useful way to summarize a column is to get the unique values it contains. This is typically most useful with character columns (e.g. species) but works with any data type.

```{r Unique}
unique(tree_data$Species)  # in this line, we are printing all of the unique species names in the dataset (in this case 2).
species_names <- unique(tree_data$Species)  # assign unique species names to a variable
```

Summarizing data is useful but we also want to be able to modify it.
So let's say we want to convert the 'Height_ft' column in `tree_data` from feet to inches. We'll start by assigning a conversion factor to the variable `in_per_ft`.

```{r InchesPerFoot}
in_per_ft <- 12  # Since there are 12 inches in a foot
```

We can then use this conversion factor to convert the 'Height_ft' column to inches, and place this converted value in a new column we will create in the data frame. Notice below that if you assign a vector to a column that doesn't exist yet, R will automatically create that column. We'll use the `head()` function to verify that the `Height_in` column was created correctly.

```{r HeightConversion}
# here we are specifying the new column on the left side of the '<-', and telling R what we want put into it on the right side of the '<-'.
tree_data$Height_in <- tree_data$Height_ft * x

# we can now use the 'head function to check our work, and make sure the proper conversion was carried out:
head(tree_data)
```