---
title: "Day 2 Module 5"
author: "Lauren Pandori & John Paul Schmit"
date: "12/15/2021"
output: html_document
---

---
#### Joining Tables {.tabset}

<details open><summary class='drop'> Joining Tables</summary>

Oftentimes we want to use data that is split into two or more data.frames. This section will go over the use of functions to "join" or combine data from two tables into one.


We will:

1) Use bind_rows() and bind_columns() for simple cases

2) Used dplyr commands for more complicated joins
  a) left_join() and right join())
  b) full join()
  c) inner_join()
  
3) Use filtering joins to filter data in one data.frame based on what is in another data.frame 
  a) semi_join()
  b) anti_join()


If you're starting with this section, run the following code chunk to get caught up. 

```{r echo=TRUE, message = FALSE, warning = FALSE}
# call packages
library('readr')
library('janitor')
library('tidyverse')

# get data
park_visits <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-09-17/national_parks.csv")
state_pop <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-09-17/state_pop.csv")
gas_price <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-09-17/gas_price.csv")

# cut some columns from park_visits data and make sure year is nuemric data only
park_visits <- park_visits %>% select(year, parkname, unit_code, state, visitors) %>% 
  filter(year!="Total") %>% mutate(year=as.integer(year))
```


</details>
<br>

<details open><summary class='drop'> Binding Rows and Columns</summary>

bind_rows() and bind_columns() simply merge two data.frames directly. 

1. bind_rows()

bind_rows() adds the rows in the second data.frame to the bottom of the first data.frame. This generally only makes sense if they have many identical columns. Make sure that columns with the same data have the same name! 

```{r eval = TRUE, include = TRUE, message = FALSE, warning = FALSE}

# we'll first make two subsets of the park_visits data
SHEN_visits <- park_visits %>%
               # get data from Shenandoah.
               filter(unit_code == 'SHEN')

SHEN_visits

ACAD_visits <- park_visits %>%
               # get data from Acadia.
               filter(unit_code == 'ACAD')

ACAD_visits

# now we will combine the 2 data.frames into one. 
SHEN_ACAD_visits <- bind_rows(SHEN_visits, ACAD_visits)

# check that data from both parks is there
SHEN_ACAD_visits %>% pull(unit_code) %>% table

rm(SHEN_ACAD_visits, SHEN_visits, ACAD_visits)
```

You generally would not split up a data.frame and then put it together again like this, but it can be useful if you have data that has the same data structure but starts out broken up by site, date, park etc. and you wish to combine it into one data.frame.

2. bind_columns()

bind_columns() adds the columns from the second data.frame to the right of the first data.frame. 


```{r eval = TRUE, include = TRUE, message = FALSE, warning = FALSE}

# we'll first make two subsets of the park_visits data
visits_left <- park_visits %>%
               # get first 2 columns.
               select(year, parkname)

visits_left

visits_right <- park_visits %>%
               #  get last 3 columns.
               select(unit_code, state, visitors)

visits_right

visits_bound<-bind_cols(visits_left, visits_right)

visits_bound

rm(visits_left, visits_right, visits_bound)
```

bind_columns() assumes that the rows are in the same order in each data.frame, and that there are no rows in one data.frame that are missing in the other. Unless you are sure that these are true, you should avoid using this function. 

</details>
<br>

<details open><summary class='drop'> bind_rows challenge </summary>

Using the park_visits data, make 2 smaller data.frames:

1) Atlantic will have just  data from Puerto Rico (PR) and the Virign Islands (VI).
2) Pacific will just have data from American Samoa (AS), Guam (GU) and Hawaii (HI)

Then, using bind_rows() combine the 2 data.frames.


</details>
<br>
<details open><summary class='drop'> bind_rows Solution </summary>

```{r eval = TRUE, include = TRUE, message = FALSE, warning = FALSE}

# Atlantic islands
Atlantic <- park_visits %>%
                     filter(state %in% c('PR', 'VI')) 
# Pacific Islands
Pacific <- park_visits %>%
                     filter(state %in% c('AS', 'GU', 'HI'))

# All islands
Islands <- bind_rows(Atlantic, Pacific)


rm(Atlantic, Pacific, Islands)

```
</details>
<br>


<details open><summary class='drop'> Basic Joins - left_join and right_join</summary>

left_join() and right_join() are a safer way to join columns from one data.frame to the columns in another.

For either of these functions, you need to specify one or more key columns using the by argument. Key columns are the data that is used to determine which rows in once data.frame match the rows in the other. The order of the rows in the data.frames does not matter.

The syntnax is left_join(x,y, by="name")   where x and y are data.frames.

The only difference between left_join and right_join is what happens when the rows in the data.frames don't match up one to one. left_join() keeps all the rows x and drops any rows in y that don't match, while right_join() keeps all the rows in y and drop any rows in x that don't match. 

Let's combine the gas_price data with the state_pop data. The gas price data does not vary by state, so we just have to match the data by year. In order to make the way these functions work a little clearer, we will first filter the state_pop data to only include data before 2000, so that there is less of an overlap between the 2 data sources. 

```{r eval = TRUE, include = TRUE, message = FALSE, warning = FALSE}
# filter the state data
state_pop2<-state_pop %>% filter(year<2000)

# left_join example 
state_pop_gas1 <-left_join(
  x=state_pop2,   #first data.frame
  y=gas_price,  # second data.frame
  by="year"  # key column
)

state_pop_gas1
```
A couple of thing to notice. The state_pop_gas1 data.frame has 5100 rows because that is what is in the state_pop2 data.frame. Every year is repeated 51 times (once for each state + DC) so the gas price data for a given year gets repeated 51 times as well. The gas_price data only goes back to 1929, so for years before that, the data just has NA for gas prices.

Here is the exact same thing, but this time using right_join

```{r eval = TRUE, include = TRUE, message = FALSE, warning = FALSE}

# left_join example 
state_pop_gas2 <-right_join(
  x=state_pop2,   #first data.frame
  y=gas_price,  # second data.frame
  by="year"  # key column
)

state_pop_gas2

```

Now the data is very different. It has 3637 rows, which seems pretty random. What has happened is:

1) All data pre-1929 was dropped
2) Data for the 71 years from 1929 to 1999 has been repeated 51 times to match the 51 states (=3621 recods)
3) The years 2000 to 2015 (15 records) each appear once as there is no matching data in the state_pop2 data.frame

So, all the data from gas_price is present, if necessary it was repeated to match up with the state_pop2 data.

</details>
<br>

<details open><summary class='drop'> Basic Joins - inner_join and full_join</summary>


In the previous section we used left and right joins to get all of the data from one data.frame with matching data from a second data.frame. Now we will look at 2 other joins:

1) inner_join() - only rows that have matches in both data frames are kept
2) full_join() - you get all the rows from both datasets even if some don't match.

inner join:
```{r eval = TRUE, include = TRUE, message = FALSE, warning = FALSE}
# inner_join example 

state_pop_gas3 <-inner_join(
  x=state_pop2,   #first data.frame
  y=gas_price,  # second data.frame
  by="year"  # key column
)

state_pop_gas3

```

<br>

Now the data.frame has 3621 rows = there are 71 years of data from 51 states. This is the smallest data.frame from these joins. The years in state_pop_gas3 are only those that are found in both data.frames. 


full join:
```{r eval = TRUE, include = TRUE, message = FALSE, warning = FALSE}
# full_join example 

state_pop_gas4 <-full_join(
  x=state_pop2,   #first data.frame
  y=gas_price,  # second data.frame
  by="year"  # key column
)

state_pop_gas4

rm(state_pop_gas1, state_pop_gas2, state_pop_gas3, state_pop_gas4)

```
This data.frame has 5116 rows, there most of all the data.frames. This includes all data frome either data.frame regardless of if there is match.

</details>
<br>


<details open><summary class='drop'> Basic Joins Challenge </summary>

Using the park_visits and the state_pop data create a new data.frame with all the data from park_visit and any matching data from state_pop.


Hint if you need to use 2 columns as keys the you write the "by=" argument like so: by=c("Column1","Column2")

</details>
<br>
<details open><summary class='drop'> Basic Join Solution </summary>

```{r eval = TRUE, include = TRUE, message = FALSE, warning = FALSE}

park_visits_pop<-left_join(park_visits, state_pop, by=c("state", "year"))

rm(park_visits_pop)

```
</details>
<br>

<details open><summary class='drop'> Filtering Joins</summary>

Sometimes you don't want to combine two tables, but you do want to filter one table so that it only has data that can match data in another table. semi_join() is used for this

full join:
```{r eval = TRUE, include = TRUE, message = FALSE, warning = FALSE}
# semi_join example 

state_pop_gas5 <-semi_join(
  x=state_pop2,   #first data.frame
  y=gas_price,  # second data.frame
  by="year"  # key column
)

state_pop_gas5

```

The columns in state_pop_gas4 are the same that are in state_pop2. However,all the data from year prior to 1929 has now been removed, because pre-1929 data is not in the gas_price data. 


anti_join is the opposite of semi_join. It will keep the data one data.frame that does not have a match in the second data.frame

full join:
```{r eval = TRUE, include = TRUE, message = FALSE, warning = FALSE}
# semi_join example 

state_pop_gas6 <-anti_join(
  x=state_pop2,   #first data.frame
  y=gas_price,  # second data.frame
  by="year"  # key column
)

state_pop_gas6

#clean up
rm(state_pop2, state_pop_gas1, state_pop_gas2, state_pop_gas3, state_pop_gas4, state_pop_gas5, state_pop_gas6)

```
state_pop_gas_6 only has 1479 rows, because it only contains data from before 1929 when the gas_price data starts. 

</details>
<br>

<details open><summary class='drop'> Filtering Joins Challenge </summary>

Filter the park_visits data so that it only contains data that can be joined with the data in and the state_pop data create a new data.frame with all the data from park_visit and any matching data from state_pop and gas_price. 

Note that the state_pop data has some NAs for population for Alaska and Hawaii prior to 1949. For a bigger challenge, don't include data those state / year combinations in your final data set.

Hint: the function is.na(data) will be TRUE when a value is NA. If you put an exclamation mark in front !is.na(data) it will tell you when the values are not NA. 

</details>
<br>
<details open><summary class='drop'> Filtering Joins Solution </summary>

```{r eval = TRUE, include = TRUE, message = FALSE, warning = FALSE}

## In this case we filter out the NA values in the semi_join function. You could also fliter them our first and save that to a new data.frame and then do the join.

park_visits_filtered<-semi_join(park_visits, state_pop %>% filter(!is.na(pop)), by=c("state", "year")) %>% 
  semi_join(gas_price, by="year")

park_visits_filtered


rm(park_visits_filtered)
```
</details>
<br>

