---
title: "Day 2 Module 4"
author: "Lauren Pandori & John Paul Schmit"
date: "12/15/2021"
output: html_document
---

---
#### Intro to Data Wrangling  - Pivoting Tables {.tabset}

<details open><summary class='drop'> Pivoting Tables</summary>

This section will go over the use of functions to change data from wide to long format and vice-versa.

We will:

1. Explain what these formats are and why anyone cares
2. pivot_longer()
3. pivot_Wider()

If you're starting with this section, run the following code chunk to get caught up. 

```{r echo=TRUE, message = FALSE, warning = FALSE}
# call packages
library('readr')
library('janitor')
library('tidyverse')

# get data
park_visits <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-09-17/national_parks.csv")
state_pop <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-09-17/state_pop.csv")
gas_price <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-09-17/gas_price.csv")

# cut some columns from park_visits data
park_visits <- park_visits %>% select(year, parkname, unit_code, state, visitors)
```

</details>
<br>
<details open><summary class='drop'>Why do we care? </summary>

Depending on your goal, you may want to have your data formatted in multiple ways. For example, it is much easier for humans to comprehend wide format data, while most R functions rely on long format data. Here is a quick demonstration using park visits data:

```{r eval = TRUE, include = TRUE, message = FALSE, warning = FALSE}

# we'll first take a subset of the park visits data
medn_visits <- park_visits %>%
               # get parks in MEDN and no totals, years 2010-2015
               filter(unit_code %in% c('CABR', 'CHIS', 'SAMO') & year != 'Total') %>%
               # make year a number, since no more text 
               mutate(year = as.numeric(year)) %>%
               # get years 2010:2015
               filter(year >= 2010 & year <=2015) %>%
               # arrange in ascending order
               arrange(year) %>%
               # select relevant columns
               select(unit_code, year, visitors)

# if we take a look at the data, it is in long format, each observation is a row
# this isn't the most human-friendly way to look at data, but it is really helpful if we want to use dplyr group_by() functions like mutate() and summarize()
medn_visits

# conversely, we can pivot longer to make the data wide format
# it is more human-friendly but you can't use dplyr functions on it easily
medn_visits_wide <- pivot_wider(medn_visits, names_from = year, values_from = visitors)
medn_visits_wide
```

There are two commands we'll discuss today:

  1. pivot_wider() : converts from long --> wide
  2. pivot_longer() : converts from wide --> long

</details>
<br>
<details open><summary class='drop'> pivot_wider()</summary>

Full disclosure, I almost always have to google the arguments that go into the pivot_ commands. Here's the website for pivot_wider() <https://tidyr.tidyverse.org/reference/pivot_wider.html>

Here's a short example - pivoting the MEDN visits from long to wide, with more detail than the prior code chunk. 

```{r eval = TRUE, include = TRUE, message = FALSE, warning = FALSE}
# check out the original data (medn_visits) in long format
medn_visits

# if we want to make each row a park and each column a year, we can pivot wider
medn_visits_wide <- medn_visits %>%
                    pivot_wider(
                                # names from is the thing you want to be the columns
                                names_from = year,
                                # values from is what you want to fill the columns with
                                values_from = visitors)

# check out the result
medn_visits_wide

# we can also make park units the columns and fill it with visitor data for each year
medn_visits_wide <- medn_visits %>%
                    pivot_wider(names_from = unit_code, values_from = visitors)
medn_visits_wide
```

</details>
<br>
<details open><summary class='drop'> pivot_longer()</summary>

Now that we have our data in wide format, it is time to revert back to pivoting longer. Here's the reference material for pivot_longer() <https://tidyr.tidyverse.org/reference/pivot_longer.html>.

```{r eval = TRUE, include = TRUE, message = FALSE, warning = FALSE}
# check out the original data (medn_visits_wide) in wide format
medn_visits_wide

# if we want to make each row an observation, with an associated park and year, we'll pivot_longer
medn_visits_long <- medn_visits_wide %>%
                    pivot_longer(
                                # first argument is the columns you'd like to transform
                                SAMO:CHIS,
                                # next is what you'd like to name the former name cols
                                names_to = 'unit_code',
                                # last is what you'd like to name the former values cols
                                values_to = 'visitors')

# check out the result
medn_visits_long

# this should look familiar - like the original medn_visits data
```

</details>
<br>
<details open><summary class='drop'> pivot challenge </summary>

This is a more involved challenge that will draw on your knowledge of both dplyr (from Module 3) and pivoting. 

Using the state_pop data, accomplish the following tasks: 

  1. Get data for 3 states of your choice for the years 2010-2015 (note - there is no 'total' values in the year column)
  2. Create a df/tibble in long format
  3. Pivot wider to make a human-friendly table with states as rows and years as columns
  4. Pivot longer, returning to the original layout
  
Hint: if your column names are numbers, like years, you must refer to them using backticks, like a column called 2020 would be `2020`

</details>
<br>
<details open><summary class='drop'> pivot solution </summary>

```{r eval = TRUE, include = TRUE, message = FALSE, warning = FALSE}
# get 3 states of choice (FL, CA, AK) for years 2010-2015
solution_original <- state_pop %>%
                     filter(state %in% c('FL', 'AK', 'CA') & 
                            year >= 2010 & year <= 2015) 
# pivot wider
solution_wide <- solution_original %>%
                 pivot_wider(names_from = 'year',
                             values_from = 'pop')

# pivot longer to return to format of solution_originol
solution_long <- solution_wide %>%
                 pivot_longer(`2010`:`2015`, names_to = 'year', values_to = 'pop')
```
</details>
<br>

