---
output: 
  html_document:
    css: custom_styles.css
title: "NPS IMD R Training I"
---

<h2>NPS Inventory and Monitoring Division R Training I: January 31 -- February 2, 2022</h2> {.tabset .tabset-pills}
--------------------------------------------------
### Prep for Training{.tabset}

```{r Install_instructions, child = "D0_Install_R_and_RStudio.Rmd"}

```

### Day 1: Intro to R {.tabset}

```{r Intro_to_R_tab, child = "D1_M1_Intro_to_R_tab.Rmd"}

```

```{r Project_setup, child = "D1_M2_Project_and_file_setup.Rmd"}

```

```{r Getting_help, child = "D1_M3_Getting_help.Rmd"}

```

```{r Types_of_data, child = "D1_M4_Types_of_Data.Rmd"}

```

```{r Data_summary_I, child = "D1_M5_Data_summary_I.Rmd"}

```

```{r Data_importing_and_subsetting, child = "D1_M6_Importing_and_Subsetting.Rmd"}

```


### Day 2: Data Wrangling {.tabset}
```{r Intro_and_Packages, child = "IMD_D2_M1_Intro_and_Packages.Rmd"}

```

```{r Data_Cleaning, child = "IMD_D2_M2_Data_Cleaning.Rmd"}

```

```{r Data_Wrangling, child = "IMD_D2_M3_Data_Wrangling.Rmd"}

```

```{r Data_Reshaping, child = "IMD_D2_M4_Pivoting_Tables.Rmd"}

```


```{r D2_Joining_Tables, child = "IMD_D2_M5_Joining_Tables.Rmd"}

```

### Day 3: Data Viz {.tabset}

```{r data_viz1, child = "IMD_D3_M1_Data_Viz.Rmd"}

```


### Day 4: Best Practices & Wrap-up {.tabset}

```{r Good_Practices, child = "D4_M1_Good_Practices.Rmd"}

```

```{r Control_Structures, child = "D4_M2_Control_Structures.Rmd"}

```

```{r Troubleshooting_Iteration, child = "D4_M3_Troubleshooting_Iteration.Rmd"}

```


### Useful websites
There's a lot of great online material for learning new applications of R. The ones I've used the most are listed below. 

<u>Online Books</u><br>
<ul>
<li><a href="https://r4ds.had.co.nz/">R for Data Science</a> First author is Hadley Wickham, the programmer behind the tidyverse. There's a lot of good stuff in here, including a chapter on using R Markdown, which is what I used to generate the website we used for training.</li>
<li><a href="https://ggplot2-book.org//">ggplot2: Elegant Graphics for Data Analysis</a> A great reference on ggplot2 also by Hadley Wickham.</li> 
<li><a href="https://bookdown.org/rdpeng/RProgDA/">Mastering Software Development in R</a> First author is Roger Peng, a Biostatistics professors at John Hopkins, who has taught a lot of undergrad/grad students how to use R. He's also one of the hosts of Not So Standard Deviations podcast. I like his intro to ggplot a lot. He's also got a lot of more advanced topics in this book, like making functions and packages.</li>
<li><a href="https://r-pkgs.org/">R Packages</a> Another book by Hadley Wickham that teaches you how to build, debug, and test R packages.</li>
<li><a href="http://adv-r.had.co.nz/">Advanced R</a> Yet another book by Hadley Wickham that helps you understand more about how R works under the hood, how it relates to other programming languages, and how to build packages.</li>
<li><a href="https://mastering-shiny.org/">Mastering Shiny</a> And another Hadley Wickham book on building shiny apps.</li>
</ul>

<u>Other useful sites</u><br>
<ul>
<li><a href="https://stat545.com/index.html">STAT545</a> Jenny Bryan's site that accompanies the graduate level stats class of the same name. I like that she includes topics on best practices for coding, and not just how to make scripts work.</li> 
<li><a href="https://rstudio.com/">RStudio home page</a> There's a ton of info in the Resources tab on this site, including cheatsheets for each package developed by RStudio (ie tidyverse packages), webinars, presentations from past RStudio Conferences, etc.</li>
<li><a href="https://support.rstudio.com/hc/en-us/articles/201057987-Quick-list-of-useful-R-packages">RStudio list of useful R packages by topic</a></li>
<li><a href="https://happygitwithr.com/">Happy Git with R</a> If you find yourself wanting to go down the path of hosting your code on github, this site will walk you through the process of linking github to RStudio.</li>
<li><a href="https://www.r-spatial.org/r/2018/10/25/ggplot2-sf.html">R-spatial mapping in ggplot</a> The r-spatial.org site is a blog with lots of interesting topics covered. The blog I linked to helped me figure out how to map shapefiles in ggplot.</li>
</ul>

### Code printout {.tabset}
```{r Code_printout, child = "IMD_Code_Printout.Rmd"}

```

### Meet the Instructors
```{r about, child = "IMD_Meet_the_Trainers.Rmd"}

```

